# Peer Assessment 1: Fitness Tracking Record
## Required packages
```{r echo = FALSE}
require(ggplot2)
require(lubridate)
require(plyr)
require(stringr)
```
## Loading and preprocessing of data
1. Read the data from the `activity.csv` file in the working directory.
- The data importation **does not exclude missing (NA) values**.
- The interval field contains time information in integer format; for example 5:30 AM is represented as 530.
- The script combines `date` and `interval` to a single POSIXct field called `timestamp`.
- Any superfluous fields are discarded.
```{r echo = TRUE}
setClass("ymd") #Now ymd is a class I can use in read.table colClasses or anywhere else
setAs("character","ymd", function(from) ymd(from))
#data frame activity potentially includes NA values
activity <- read.table("activity.csv", header = TRUE, sep = ",", colClasses = c("integer", "ymd", "integer"), stringsAsFactors = FALSE)
activity$timestamp <- parse_date_time(paste(activity$date, str_pad(activity$interval, 4, pad = "0")), "%Y%m%d %H%M")
activity$date <- NULL
```
## Miscellaneous R functions
```{r}
formatInterval <- function(interval) {
  # Formats the interval type from the original dataset as "HH:MM" (hours and minutes)
  temp <- str_pad(interval, 4, pad = "0")
  paste(substr(temp, 1, 2), substr(temp, 3, 4), sep = ":")
}

formatTimePosix <- function(timestamp) {
  # Designed for POSIXct but may work with other datetime types, extracts time portion of POSIX
  # Formats the character return type as "HH:MM" (hours and minutes)
  paste(str_pad(hour(timestamp), 2, pad = "0"), str_pad(minute(timestamp), 2, pad = "0"), sep = ":")
}
```
### Structure of activity data frame
```{r}
str(activity)
```
## What is the mean total number of steps taken per day?
NA values omitted in this section
### Histogram of Total Daily Steps:
```{r fig.width=10, fig.height=6}
dailySteps <- ddply(na.omit(activity), "as.Date(timestamp)", summarize, sum(steps))
colnames(dailySteps)[2] <- "steps"
qplot(steps, data = dailySteps, main = "Histogram of Total Daily Steps", xlab = "steps")
```
### Mean of daily steps:
```{r}
mean(dailySteps$steps)
```
### Median of Steps each day:
```{r}
median(dailySteps$steps)
```
## What is the average daily pattern?
* It was unclear from the requirements if NA values should be omitted for this part.  
* Since the requirements do not mention imputing missing values until the next section, the author chose to omit NA values.
* This code chunk calculates the mean of steps aggregated by interval (time of day), irrespective of the specific day.
* Has smoothing line for reference

### Possible artifacting in plots vs Interval
Interval is an interval field encoding the time. The last two digits represent the minutes; therefore these digits can have values only between 00 and 59. The missing values between 60 and 99 may distort the graph shape on very small scales; however even if this is the case the effect might not be noticable.
```{r fig.width=10, fig.height=6}
timeOfDaySteps <- ddply(na.omit(activity), "interval", summarize, mean(steps))
colnames(timeOfDaySteps)[2] <- "meanOfSteps"
qplot(interval, meanOfSteps, data = timeOfDaySteps, geom = c("line", "smooth"), main = "Mean steps vs time of day", xlab = "Interval (time of day)", ylab = "mean of steps") #equivalent of plot(x, y, type = "l", ...) with smoothing added (geom = "smooth")
```
### Which 5 minute interval (averaging all days) contains the maximum number of steps?
The Top 5 means of steps are shown for comparison.
```{r}
stepsMaxPerTime <- head(timeOfDaySteps[ order(-timeOfDaySteps$meanOfSteps), ])
stepsMaxPerTime$time = formatInterval(stepsMaxPerTime$interval)
stepsMaxPerTime$interval <- NULL
stepsMaxPerTime[c(2,1)] #display time first
```
## Imputing missing values
### Count of pre-imputation activity dataset missing values
```{r}
sum(is.na(activity$steps))
```
### Fill in missing values (imputation)
Replace missing (NA) values with the mean of steps for the same time period (on other days).  
The new `activity.imputed` dataset contains the imputed values. Superfluous datasets are dropped.
```{r}
activity.na <- activity[is.na(activity$steps), c("interval", "timestamp")]
activity.imputed <- merge(activity.na, timeOfDaySteps, by = "interval")
rm(activity.na)
colnames(activity.imputed)[3] <- "steps"
activity.imputed <- rbind(activity.imputed, na.omit(activity))
rm(activity)
```
### Count of post-imputation activity dataset missing values
```{r}
sum(is.na(activity.imputed$steps))
```
### Structure of post-imputation activity dataset
```{r}
str(activity.imputed)
```
### What is the mean total number of steps taken per day?
* Using the `activity.imputed` dataset, this section repeats the previous analysis for comparison.  
* The imputation substantially increased the step frequency amplitudes though the overall shape is similar to the pre-imputation histogram.

#### Histogram of Total Daily Steps:
```{r fig.width=10, fig.height=6}
dailySteps <- ddply(activity.imputed, "as.Date(timestamp)", summarize, sum(steps))
colnames(dailySteps)[2] <- "steps"
qplot(steps, data = dailySteps, main = "Histogram of Imputed Total Daily Steps", xlab = "steps")
```
#### Mean of daily steps:
```{r}
mean(dailySteps$steps)
```
#### Median of Steps each day:
```{r}
median(dailySteps$steps)
```
## Differences in activity level between weekdays and weekend
* Uses the `activity.imputed` data set.
* Author anticipates it should run on non-English localized systems; it refers to weekdays by numeric index rather than name.
* Create new `dayType` factor in the dataset which can be either "weekday" or "weekend"
```{r}
#wday() outputs numeric weekday, avoiding any localization issues. Day 1 is Sunday, 2 is Monday...
activity.imputed$dayType <- factor(ifelse(wday(activity.imputed$timestamp) %in% 2:6, "weekday", "weekend"), levels = c("weekend", "weekday")) #first factor is weekend, for plot panel order
str(activity.imputed)
meanStepsByTimeAndType <- ddply(activity.imputed, c("dayType", "interval"), summarize, mean(steps))
colnames(meanStepsByTimeAndType)[3] <- "meanOfSteps"
```
### Sample of dataset used for mean of steps by time and day type (weekend vs weekday)
```{r}
head(meanStepsByTimeAndType)
```
### Plot of mean of Steps vs interval (time of day), separate panels for weekend vs weekday
```{r fig.width=10, fig.height=6}
qplot(interval, meanOfSteps, data = meanStepsByTimeAndType, color = I("blue"), geom = "line", facets = dayType ~ ., xlab = "Interval", ylab = "Number of steps")
```